# Explanatory Data Analysis
## Introduction to EDA
**Exploratory Data Analysis (EDA)** is a critical process in the data analysis workflow. It involves examining and visualizing a dataset to uncover patterns, spot anomalies, test hypotheses, and check assumptions using summary statistics and graphical representations. EDA provides a solid understanding of the data and lays the foundation for more advanced statistical analyses or machine learning models.

The key objectives of EDA are;- 

1. Understand and familiarize the data structure.
2. Detect outliers and anomalies.
3. Identify patterns and trends. 
4. Check and verify assumptions, etc. 

By performing EDA thoroughly, the groundwork is laid for more accurate results, ultimately provides insights that lead to better decision-making. EDA also guides to better model-selection. 

Here are the common techniques involved in EDA;- 

- **Summary Statistics**: The analysts compute the summary statistics like mean, median, mode and standard deviation to find the spread and central tendency in the data. 
- **Data Visualization**: Charts and graphs are drawn to provide a visual interpretation of the data set. 
- **Data Cleaning and Preparation**: Data manipulation and pre-processing is done to reduce the risk of misinterpretation.

There are more EDA techniques like univariate, bivariate and multivariate analysis, and inferential statistics. 

## Choosing a data set 
You are required to select a data set from the list below and perform EDA;

- MBA Admission dataset, Class 2025 - [download here](https://www.kaggle.com/datasets/taweilo/mba-admission-dataset)
- Global Black Money Transactions Dataset - [download here](https://www.kaggle.com/datasets/waqi786/global-black-money-transactions-dataset)
- Crop Yield Prediction Dataset - [download here](https://www.kaggle.com/datasets/mrigaankjaswal/crop-yield-prediction-dataset)
- ChickWeight inbuilt R dataset - load from the command `data("ChickWeight")`
- Seatbelts data set - load from the command `data("Seatbelts")`
- The "Groceries" from the R package  comes  `arules`- Load the data by first importing arules(`library(arules)`) then the data by `data("Groceries")`
- `"CreditCard"` data from package `AER`- Load the data by first importing AER (`library(AER)`) then the data by `data("CreditCard")`

Outline the questions to explore during EDA based on the Data set you have chosen

## Conducting EDA
You will use the skills acquired in this camp(especially `dplyr` and `ggplot2`) to;

- Explore the data set structure 
- Visualize the relationship between variables 
- Summarize key findings 

## Present Findings 
Finally, you will prepare a brief presentation of your findings ,and discuss on the insights obtained and the techniques used. 

_______________________________________________________________________
## <span style="color: brown;">EDA - Example</span> 

This is just an example on how to explore EDA. Download the Ecommerce data set from [here](https://www.kaggle.com/datasets/shrishtimanja/ecommerce-dataset-for-data-analysis/data) 

### <span style="color: brown;">Introduction</span> 

This eCommerce dataset contains transaction-level information, including customer demographics (e.g., age, gender), purchase details (e.g., product category, payment method), and discount information. It captures both gross and net amounts, helping to analyze the impact of discounts and customer behavior across different locations. This data set provides insights into customer spending patterns and promotional effectiveness

### <span style="color: brown;">Data Assessment and Cleaning</span> 

Load the necessary data sets and libraries 
```{r}
# Load the libraries 
library(dplyr)
library(ggplot2)

# Load the data 
ecommerce <- read.csv("data/ecommerce.csv")
```

Show the first five rows 
```{r}
head(ecommerce)
```

The data set features/columns and the shape of the data set 
```{r}
colnames(ecommerce) #the data set features 

dim(ecommerce) # the shape of the data
```
There are 55000 observations and 13 features/columns in the data set. Lets now find the data types of the columns'

```{r}
str(df)
```

There are 5 numeric columns and 8 non numeric columns. 

Find and count the null values in the data set 
```{r}
sum(is.na(ecommerce))
```

The data set is complete with no missing values. Lets now find if there are any duplicated records 
```{r}
sum(duplicated(ecommerce))
```

There are no duplicated records

Calculate the summary statistics 
```{r}
summary(ecommerce)
```

### <span style="color: brown;">Data Visualization</span> 

The relationship between the discount amount and the gross amount using a scatter plot

```{r}
# Create a scatter plot 
ggplot(ecommerce, # data
       aes(x = Gross.Amount, y = Discount.Amount..INR.)) + # aesthetics
  geom_point() + 
  labs(
    title="The relationship between Gross Amount and Discount",
    y="Discount Amount",
    x="Gross Amount"
  )
```

There is no clear relationship between the discount amount and the gross amount 

The average discount per product category
```{r}
discount_per_category <- ecommerce %>%
  select(Product.Category, Discount.Amount..INR.) %>%
  group_by(Product.Category) %>%
  summarise(Average.Discount=mean(Discount.Amount..INR.))

discount_per_category
```

Plot the data on a bar chart 
```{r}
# Create a bar chart 
ggplot(discount_per_category, aes(x = Product.Category, y = Average.Discount)) +
  geom_bar(stat = "identity") +
  labs(title = "Average discount per the product category", 
       x = "Product Category", 
       y = "Average Discount") +
  theme_classic()
```

The average discount was almost equal for all the categories 

Count the purchases by purchase method
```{r}
purchase_method_count <- ecommerce %>% 
  group_by(Purchase.Method)%>%
  summarize(Count=n())

purchase_method_count
```
Display on a bar chart

```{r}
# Create a bar chart 
ggplot(purchase_method_count, aes(x = Purchase.Method, y = Count)) +
  geom_bar(stat = "identity") +
  labs(title = "Count of purchases by purchase method", 
       x = "Purchase Method", 
       y = "Count") +
  theme_classic()
```

Most customers purchased goods by credit and debit cards. 

Show the net amount distribution by histogram 
```{r}
# Create a histogram
ggplot(ecommerce, aes(x = Net.Amount)) +
  geom_histogram(bins = 8, fill = "blue", color = "black") +
  labs(title = "Distribution of Net Amount", 
       x = "Net Amount", 
       y = "frequency") +
  theme_minimal()
```
Most products range from 1000 to 4000 Indian Rupees

Find out the different age groups 
```{r}
unique(ecommerce$Age.Group)
```

There are five different age groups, lets inspect their spending characteristics on the eCommerce site using box plots 
```{r}
# Plot a box plot 
ggplot(ecommerce, 
       aes(x=Age.Group, y=Gross.Amount, fill = Age.Group))+
  geom_boxplot() +
  labs(title="Spending based on different age groups",
       x = "Age groups",
       y = "Gross Amount") +
  theme_classic()
```

There was no clear difference between the spending patterns based on the age groups 

### <span style="color: brown;">Key Findings</span> 

These were the key findings that were discovered;

- The amount spend on purchase and the category of product purchased had no effect on the discount given.
- Most of the customers find it easy using their credit and debit cards. This may be due to low fees imposed , transaction speeds, not adopted alternatives methods, or the popularity of these cards when purchasing.
- There was no clear difference between the spending patterns based on the age groups. 

<span style="color: brown;">**________________________________________________________________________________**</span>