# Capstone Project
## Introduction
### Project Overview and Objectives 
- The main objective of this capstone project is to allow the student apply all the skills and techniques learned through the camp in a comprehensive real-world analysis.

- It greatly emphasizes on teamwork, critical thinking and creativity

### Group Formation
Participants are divided into 5 groups, each consisting of 5 members.
Each group is assigned a different data set. These are the  proposed data sets and their links to download them:


1. **Healthcare Data set**

https://www.kaggle.com/datasets/imtkaggleteam/mental-health?select=1-+mental-illnesses-prevalence.csv

2. **Retail Sale data**

https://www.kaggle.com/datasets/manjeetsingh/retaildataset

3. **Financial Transaction Data**

https://www.kaggle.com/datasets/apoorvwatsky/bank-transaction-data

https://www.kaggle.com/datasets/mlg-ulb/creditcardfraud

4. **Social Media Data**

https://www.kaggle.com/datasets/datasnaek/youtube-new?select=USvideos.csv

https://www.kaggle.com/datasets/datasnaek/youtube-new?select=GBvideos.csv

5. **Environmental Data set**

https://www.kaggle.com/datasets/sevgisarac/temperature-change

## Project Planning
### Defining the problem statement
Using your assigned data set, you will be required to;

- define a clear problem statement to address. 
- narrow down on the scope of the project to make it manageable within the given timeframe
- outline your project goals and the expected outcomes 


### Planning the Analysis WorkFlow 
In your groups, you are required to plan the analysis including:

* **Data Cleaning and Preprocessing**

You will identify potential issues for instance missing values and outliers ,then plan how to address them 

* **Explanatory Data Analysis**

Plan how to conduct the initial exploration of the data set to understand the data structure and key patterns

* **Model Selection and Application**

Discuss on the potential statistical or machine learning models that could be applied to the data 

* **Evaluation and Interpretation**

Plan how to evaluate the model’s performance and interpret the results.


## Project Execution
### Part 1
#### Data Cleaning and Preprocessing 
The groups will begin by cleaning and preparing the data sets including;

- Handling missing values
- Correcting inconsistencies in the data set
- Ensuring data is ready for analysis 

The instructirs will provide support and guidance as needed, answering questions and help in troubleshooting issues

#### Explanatory Data Analysis(EDA)

The groups will then;

- Perform EDA to uncover insights, identify trends and explore the relationships between variables 
- Visualize the data using `ggplot2` or any other tool to better understand the data.
- Document their findings and decide on the next step in the analysis 

### Part 2
#### Model Building and Analysis 
Each group is required to; 

- Select and apply an appropriate statistical or machine learning model to address their research questions 
- Train, test and fine-tune the model as needed. 
- Emphasize on the interpreting the model outputs, understanding the implications and ensuring the results align with the project goals 

#### Evaluation and Insights
The groups will then; 

- evaluate the performance of their models using relevant metrics (e.g., accuracy, RMSE, precision, recall).
- discuss on the insights gained from the analysis and how these insights could be applied in a real-world context.
- prepare visualizations and summary reports to effectively communicate their findings.

## Project Presentations and Feedback
Each group will finally present their project that will include; 

- An overview of the data set and the problem statement.
- The steps taken during data cleaning, EDA, and model building.
- The key findings, model performance, and final insights.
- Visualizations and any actionable recommendations based on the analysis.

_______________________________________________________________________
## <span style="color: brown;">Example: Ecommerce Project</span>

Load the libraries
```{r}
library(dplyr)
library(ggplot2)
library(caret)
```

Load the data
```{r}
ecommerce <- read.csv("data/ecommerce.csv")

# Get a glimpse of the data set
str(ecommerce)
```

### <span style="color: brown;">Part 1</span>
<span style="color: brown;">**Data Cleaning and Preprocessing**</span>

We check for missing values in the data set to understand if any imputation or removal is needed. The columns that have missing values so we can decide whether to impute or remove them
```{r}
# Count of missing values per column
sapply(ecommerce, function(x) sum(is.na(x)))
```
There were no null values in the data set.

The outliers in the `Gross.Amount` and `Net.Amount` columns were then identified using IQR method. The removal of outliers ensures a more reliable analysis, particularly for numerical data. 
```{r}
# Outlier detection function using IQR
identify_outliers_iqr <- function(column) {
  Q1 <- quantile(column, 0.25, na.rm = TRUE)
  Q3 <- quantile(column, 0.75, na.rm = TRUE)
  IQR_value <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR_value
  upper_bound <- Q3 + 1.5 * IQR_value
  return(column < lower_bound | column > upper_bound)
}

# Removing outliers
outliers <- identify_outliers_iqr(ecommerce$Net.Amount)
ecommerce_clean <- ecommerce[!outliers, ]
```

Shape of the old ecommerce data set
```{r}
dim(ecommerce)
```

Shape of the new ecommerce data set 
```{r}
dim(ecommerce_clean)
```

There was only one outlier in the entire dataset that was identified and removed. 

The `Purchase.Date` column needed to be converted to a proper date column 
```{r}
# Convert 'Purchase Date' to Date type
ecommerce_clean$Purchase.Date <- as.Date(ecommerce_clean$Purchase.Date, format="%Y-%m-%d")

# Confirm the operation
str(dplyr::select(ecommerce_clean, Purchase.Date))
```

<span style="color: brown;">**Explanatory Data Analysis**</span>

Explanatory data analysis was performed to understand tge distribution, relationships and the underlying structure of the data set. 

The distribution of Gross Amount was plotted 
```{r}
# Histogram for Gross Amount
ggplot(ecommerce_clean, aes(x = Gross.Amount)) +
  geom_histogram(bins = 20, fill = "blue", color = "black") +
  labs(title = "Distribution of Gross Amount", x = "Gross Amount (INR)", y = "Frequency") +
  theme_minimal()
```

This histogram shows that the data appears to be right-skewed, meaning most transactions occur within the lower range (around 1000-4000 INR), with fewer transactions as the Gross Amount increases. There is a gradual decline in frequency for higher amounts, with very few transactions beyond 6000 INR. The highest frequency is concentrated between 1000 and 4000 INR. 

The bar chart to compare the number of transactions across different of product categories in the store was plotted 
```{r}
# Bar chart for Product Category
ggplot(ecommerce_clean, aes(x = Product.Category)) +
  geom_bar(fill = "blue") +
  labs(title = "Number of Transactions by Product Category", x = "Product Category", y = "Transaction Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The bar chart shows the number of transactions by product category. Electronics has the highest number of transactions, followed by Clothing and Beauty and Health. Categories like Books, Home & Kitchen, and Other have moderate transaction counts, while categories like Pet Care, Toys & Games, and Sports & Fitness have significantly fewer transactions. The chart highlights which product categories are more popular in terms of transactions

Finally, the distribution of Net Amount spent by gender was plotted on a boxplot.
```{r}
# Box plot of Net Amount by Gender
ggplot(ecommerce_clean, aes(x = Gender, y = Net.Amount)) +
  geom_boxplot(fill = c("pink", "lightblue", "gray")) +
  labs(title = "Net Amount by Gender", x = "Gender", y = "Net Amount (INR)") +
  theme_minimal()
```

This boxplot illustrates the distribution of the Net Amount (INR) spent by customers, categorized by Gender (Female, Male, Other). Here's a brief description:

- **Female**: The median net amount spent is slightly above 3000 INR, with most of the data falling between 2000 and 4500 INR. There are a few outliers above 8000 INR.
- **Male**: The median is close to that of females, slightly below 4000 INR. The interquartile range (IQR) is similar, and there are also outliers.
- **Other**: The distribution follows a similar pattern, with the median around 4000 INR. The spread of data is broader, and outliers are visible above 8000 INR.

The boxplot suggests that, overall, net spending is relatively consistent across genders, with some variations in the range of outliers

### <span style="color: brown;">Part 2</span>
<span style="color: brown;">**Model Building and Analysis**</span>

In this project, we agreed upon building a multiple linear regression model where the `Net.Amount` was the target variable and the independent (predictor) variables will include the features that might influence the `Net.Amount`, such as; `Age.Group`, `Gender`, `Discount.Availed`, `Purchase.Method`, `Product.Category` and `Location`.

Categorical variables like Gender, Age Group, and Purchase Method were then converted into numerical formats. 
```{r}
# Convert categorical variables to factors
ecommerce_clean$Gender <- as.factor(ecommerce_clean$Gender)
ecommerce_clean$Age.Group <- as.factor(ecommerce_clean$Age.Group)
ecommerce_clean$Purchase.Method <- as.factor(ecommerce_clean$Purchase.Method)
ecommerce_clean$Product.Category <- as.factor(ecommerce_clean$Product.Category)
ecommerce_clean$Location <- as.factor(ecommerce_clean$Location)
```

The data was then split into train and test data sets 
```{r}
# Split the data into training (70%) and testing (30%) sets
set.seed(123)  # For reproducibility
sample_split <- sample(1:nrow(ecommerce_clean), 0.7 * nrow(ecommerce_clean))
train_data <- ecommerce_clean[sample_split, ]
test_data <- ecommerce_clean[-sample_split, ]
```

The categorical variables converted to numerical were then encoded.
```{r}
# Encode categorical variables in the training set
train_data_encoded <- model.matrix(Net.Amount ~ Gender + Age.Group + Purchase.Method + Discount.Availed + Product.Category + Location, data = train_data)
train_data_encoded <- as.data.frame(train_data_encoded)

# Keep the Net.Amount column in train set
train_data_encoded$Net.Amount <- train_data$Net.Amount

# Repeat for the test data 
# Encode categorical variables in the test set
test_data_encoded <- model.matrix(Net.Amount ~ Gender + Age.Group + Purchase.Method + Discount.Availed + Product.Category + Location, data = test_data)
test_data_encoded <- as.data.frame(test_data_encoded)

# Keep the Net.Amount column in test set
test_data_encoded$Net.Amount <- test_data$Net.Amount
```

Build the linear regression model
```{r}
# Build the multiple linear regression model on the training data
model <- lm(Net.Amount ~ ., data = train_data_encoded)

# Summary of the model
summary(model)
```

Predictions were then performed by the model
```{r}
# Predict Net Amount on test data
predictions <- predict(model, newdata = test_data_encoded)

# View the first few predictions
head(predictions)
```

<span style="color: brown;">**Evaluation and insights**</span>

The model was evaluated using Mean Squared Error(MSE) and Root Squared Error(RMSE)
```{r}
# Calculate predictions on the test set
test_predictions <- predict(model, newdata = test_data_encoded)

# Calculate MSE and RMSE
mse <- mean((test_predictions - test_data_encoded$Net.Amount)^2)
rmse <- sqrt(mse)

cat("Mean Squared Error (MSE):", mse, "\n")
cat("Root Mean Squared Error (RMSE):", rmse, "\n")
```

A QQ plot was used to assess if the residuals follow a normal distribution.

```{r}
# QQ Plot
qqnorm(model$residuals)
qqline(model$residuals, col = "red")
```

The predicted values have a different pattern(straight-line) with the original values(sigmoid-curve)

### <span style="color: brown;">Presentation and Feedback</span>

The model identifies some statistically significant predictors, its overall fit is quite average, as indicated by unexpected R-squared values and substantial error metrics (MSE and RMSE). The limited explanatory power suggests that other relevant variables or non-linear relationships may exist and could be explored to improve the model's performance. Further diagnostics, feature engineering, or transformations might be necessary to enhance predictive accuracy

<span style="color: brown;">**________________________________________________________________________________**</span>